---
title: "NESCAC Position Player WAR and wRC"
author: "Tyler Marshall"
date: "12/23/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages Used

```{r}
library(tidyverse)
library(robotstxt)
library(knitr)
library(rvest)
```


## Web Scraping

First, I want to make sure that the NESCAC website allows us to scrape data from it. In this case it does, so we can continue by scraping what we need. I have to scrape the data for each team individually and will bind it together in one dataset.
```{r}
paths_allowed("https://www.nescac.com/sports/bsb/2018-19/players?sort=ab&view=&pos=h&r=0")
```


### Amherst
```{r}

amherst_page <- read_html("http://athletics.amherst.edu/sports/bsb/2018-19/teams/amherst?view=lineup&r=0&pos=") 
#URL for Amherst stats

amherst_tables <- html_nodes(amherst_page, "table") #Read in data
amherst <- html_table(amherst_tables[[4]]) #Hitting Data
amherst2 <- html_table(amherst_tables[[5]]) #Extended Hitting Data

amhersthit <- inner_join(amherst, amherst2, by="Name") #Join two hitting datasets
names(amhersthit)[9]<-"Doubles" #Naming column Doubles
names(amhersthit)[10]<-"Triples" #Naming column Triples

amhersthit2 <- amhersthit %>%
  mutate(School= "Amherst") #Add school to be used later
```


### Bates
```{r}
bates_page <- read_html("https://www.gobatesbobcats.com/sports/bsb/2018-19/teams/bates") 
#URL for Bates data

bates_tables <- html_nodes(bates_page, "table") #Read in data
bates <- html_table(bates_tables[[4]]) #Hitting Data
bates2 <- html_table(bates_tables[[5]]) #Extended Hitting data

bateshit <- inner_join(bates, bates2, by="Name") #Join two hitting datasets
names(bateshit)[9]<-"Doubles" #Naming column Doubles
names(bateshit)[10]<-"Triples" #Naming column Triples

bateshit2 <- bateshit %>%
  mutate(School= "Bates") #Add school to be used later
```

### Bowdoin
```{r}

bowdoin_page <- read_html("https://athletics.bowdoin.edu/sports/bsb/2018-19/teams/bowdoin") 
#URL for Bowdoin data

bowdoin_tables <- html_nodes(bowdoin_page, "table") #Read in data
bowdoin <- html_table(bowdoin_tables[[4]]) #Hitting data
bowdoin2 <- html_table(bowdoin_tables[[5]]) #Extended hitting data

bowdoinhit <- inner_join(bowdoin, bowdoin2, by="Name") #Join two hitting datasets
names(bowdoinhit)[9]<-"Doubles" #Naming column Doubles
names(bowdoinhit)[10]<-"Triples" #Naming column Triples

bowdoinhit2 <- bowdoinhit %>%
  mutate(School= "Bowdoin") #Add school to be used later
```

### Colby
```{r}
colby_page <- read_html("https://gocolbymules.com/sports/bsb/2018-19/teams/colby") 
#URL for data

colby_tables <- html_nodes(colby_page, "table") #Read in data
colby <- html_table(colby_tables[[4]]) #Hitting data
colby2 <- html_table(colby_tables[[5]]) #Extended hitting data

colbyhit <- inner_join(colby, colby2, by="Name") #Join two hitting datasets
names(colbyhit)[9]<-"Doubles" #Naming column Doubles
names(colbyhit)[10]<-"Triples" #Naming column Triples

colbyhit2 <- colbyhit %>%
  mutate(School= "Colby") #Add school to be used later
```

### Hamilton
```{r}
hamilton_page <- read_html("https://athletics.hamilton.edu/sports/bsb/2018-19/teams/hamilton") 
#URL for Hamilton data

hamilton_tables <- html_nodes(hamilton_page, "table") #Read in data
hamilton <- html_table(hamilton_tables[[4]]) #Hitting data
hamilton2 <- html_table(hamilton_tables[[5]]) #Extended hitting data

hamiltonhit <- inner_join(hamilton, hamilton2, by="Name") #Join two hitting datasets
names(hamiltonhit)[9]<-"Doubles" #Naming column Doubles
names(hamiltonhit)[10]<-"Triples" #Naming column Triples

hamiltonhit2 <- hamiltonhit %>%
  mutate(School="Hamilton") #Add school to be used later

```


### Middlebury
```{r}
midd_page <- read_html("https://athletics.middlebury.edu/sports/bsb/2018-19/teams/middlebury") 
#URL for Middlebury data

midd_tables <- html_nodes(midd_page, "table") #Read in data
middlebury <- html_table(midd_tables[[5]]) #Hitting data
middlebury2 <- html_table(midd_tables[[7]]) #Extended hitting data

middleburyhit <- inner_join(middlebury, middlebury2, by="Name") #Join two hitting datasets
names(middleburyhit)[9]<-"Doubles" #Naming column Doubles
names(middleburyhit)[10]<-"Triples" #Naming column Triples

middleburyhit2 <- middleburyhit %>%
  mutate(School= "Middlebury") #Add school to be used later
```

### Trinity
```{r}
trinity_page <- read_html("https://www.bantamsports.com/sports/bsb/2018-19/teams/trinityconn") 
#URL for Trinity data

trinity_tables <- html_nodes(trinity_page, "table") #Read in data
trinity <- html_table(trinity_tables[[4]]) #Hitting data
trinity2 <- html_table(trinity_tables[[5]]) #Extended hitting data

trinityhit <- inner_join(trinity, trinity2, by="Name") #Join two hitting datasets
names(trinityhit)[9]<-"Doubles" #Naming column Doubles
names(trinityhit)[10]<-"Triples" #Naming column Triples

trinityhit2 <- trinityhit %>%
  mutate(School= "Trinity") #Add school to be used later
```

### Tufts
```{r}
tufts_page <- read_html("https://www.gotuftsjumbos.com/sports/bsb/2018-19/teams/tufts")
#URL for Tufts data

tufts_tables <- html_nodes(tufts_page, "table") #Read in data
tufts <- html_table(tufts_tables[[4]]) #Hitting data
tufts2 <- html_table(tufts_tables[[5]]) #Extended hitting data

tuftshit <- inner_join(tufts, tufts2, by="Name") #Join two hitting datasets
names(tuftshit)[9]<-"Doubles" #Naming column Doubles
names(tuftshit)[10]<-"Triples" #Naming column Triples

tuftshit2 <- tuftshit %>%
  mutate(School= "Tufts") #Add school to be used later

```


### Wesleyan
```{r}
wesleyan_page <- read_html("https://athletics.wesleyan.edu/sports/bsb/2018-19/teams/wesleyanconn") 
#URL for Wesleyan data

wesleyan_tables <- html_nodes(wesleyan_page, "table") #Read in data
wesleyan <- html_table(wesleyan_tables[[4]]) #Hitting data
wesleyan2 <- html_table(wesleyan_tables[[5]]) #Extended hitting data

wesleyanhit <- inner_join(wesleyan, wesleyan2, by="Name") #Join two hitting datasets
names(wesleyanhit)[9]<-"Doubles" #Naming column Doubles
names(wesleyanhit)[10]<-"Triples" #Naming column Triples

wesleyanhit2 <- wesleyanhit %>%
  mutate(School= "Wesleyan") #Add school to be used later
```

### Williams
```{r}
williams_page <- read_html("https://ephsports.williams.edu/sports/bsb/2018-19/teams/williams") 
#URL for Williams data

williams_tables <- html_nodes(williams_page, "table") #Read in data
williams <- html_table(williams_tables[[4]]) #Hitting data
williams2 <- html_table(williams_tables[[5]]) #Extended hitting data

williamshit <- inner_join(williams, williams2, by="Name") #Join two hitting datasets
names(williamshit)[9]<-"Doubles" #Naming column Doubles
names(williamshit)[10]<-"Triples" #Naming column Triples

williamshit2 <- williamshit %>%
  mutate(School= "Williams") #Add school to be used later
```

## Binding Datasets
```{r}

nescachit <- rbind(amhersthit2, bateshit2, bowdoinhit2, colbyhit2, hamiltonhit2, 
                   middleburyhit2, trinityhit2, tuftshit2, wesleyanhit2, williamshit2) 
#Bind each NESCAC school hitting data into one dataset
```

## Cleaning Up the Data

This step cleans up the data, so it is in the proper format to allow the analysis to be done. 
```{r}
nescachit <- nescachit %>%
  filter(Name != "Totals") %>% #Remove Totals
  filter(Name != "Opponent") %>% #Remove Opponents
  mutate(Doubles= as.numeric(Doubles)) %>% #Make numeric variables numeric
  mutate(Triples= as.numeric(Triples)) %>%
  mutate(hr= as.numeric(hr)) %>%
  mutate(hbp= as.numeric(hbp)) %>%
  mutate(sf= as.numeric(sf)) %>%
  mutate(sh= as.numeric(sh)) %>%
  mutate(Singles= h-Doubles-Triples-hr) %>% #Create a variable for Singles
  mutate(bb= as.numeric(bb)) %>%
  mutate(obp= as.numeric(obp)) %>%
  mutate(sb= as.numeric(sb)) %>%
  mutate(cs= as.numeric(cs)) %>%
  mutate(slg= as.numeric(slg))
nescachit[is.na(nescachit)] <- 0 #Replace all NAs with 0s

```

## Calculations

In the following section, all of the calculations for the necessary statistics are made. The preliminary calculations are done to determine the coefficent for wOBA and wOBA scale for the NESCAC in 2019. Also, lgwSB is calculated which is necessary for the wSB calculation. Runs Per Win is calculated to determine how many runs was worth one win in the NESCAC in 2019 and runs per plate appearence was calculated to be used in the wRC calculation.

Once the necessary preliminary calculations were made OPS, wOBA, wRAA, wSB, fOWar and wRC were all calculated for each individual player. A Shiny App was then created to display the results of this project.
```{r}

nescactotal <- nescachit %>%
  mutate(totalbb= sum(bb)) %>% #Finding totals to be used to determine wOBA coefficient
  mutate(totalhbp = sum(hbp)) %>%
  mutate(totalsingles= sum(Singles)) %>%
  mutate(totaldoubles= sum(Doubles)) %>%
  mutate(totaltriples= sum(Triples)) %>%
  mutate(totalhr= sum(hr)) %>%
  mutate(totalpa= sum(ab + bb + sf + hbp)) %>%
  mutate(totalsb= sum(sb)) %>%
  mutate(totalcs= sum(cs)) %>%
  mutate(totalruns= sum(r)) %>%
  mutate(averageobp= sum(obp)/195) %>% #Finding average OBP since average wOBA equals average OBP
  mutate(coef = (.55*totalbb + .57 * totalhbp + .7 * totalsingles + 
                   totaldoubles + 1.27*totaltriples + 
                   1.65* totalhr)/(totalpa)) %>% #Coefficient for wOBA
  mutate(wOBAscale= averageobp/coef) %>% #Found wOBAscale for NESCAC
  mutate(lgwSB= (totalsb *.2 + totalcs * -.4)/(totalsingles+totalbb+totalhbp)) %>% 
  #Found lgwSB for NESCAC using .2 for runSB and -.4 for runCS (based on MLB recent years)
  mutate(RPW = 9 *(totalruns/2871.6)*1.5 +3) %>% 
  #2871.6 is total innings, calculating Runs per Win
  mutate(RperPA = totalruns/totalpa) #Calculate runs per plate appearance needed for wRC

nescacwar <- nescachit %>%
  mutate(OPS= obp+slg) %>%
  mutate(wOBA = ((1.29*.55)*bb + (1.29*.57) *hbp + (1.29*.7)*Singles 
                 + (1.29) * Doubles + (1.29*1.2) * Triples 
                 + (1.29*1.65)* hr)/ (ab+bb+hbp+sf)) %>%
  mutate(wRAA= ((wOBA- .356)/1.29)*(ab+bb+hbp+sf+sh)) %>%
  mutate(wSB= (sb*.2) + (cs*-.4) - (.0176 * (Singles+bb+hbp))) %>%
  mutate(fOWar= (wRAA + wSB + 20/600 * pa)/ 13.36) %>% 
  #Using Fangraphs War (only offense), divided by runs per win which is 13.36 for the NESCAC
  mutate(wRC= (((wOBA- .356/1.29)) + .16)*pa) #wRC Calculation

#Add in Class Year to be used in Shiny App
nescacwar2 <- nescacwar %>%
  mutate(Year = Yr.x) %>%
  mutate(Year= case_when(grepl("Sr", Year) ~ "Senior",
         grepl("Jr", Year) ~ "Junior",
         grepl("So", Year) ~ "Sophomore",
         grepl("Fr", Year) ~ "First-Year",
         grepl("Fy", Year) ~ "First-Year",
         grepl("19", Year) ~ "Senior",
         grepl("20", Year) ~ "Junior",
         grepl("21", Year) ~ "Sophomore",
         grepl("22", Year) ~ "First-Year",
         grepl("Christian  Beal", Name) ~ "Sophomore",
         grepl("Giovanni  Torres", Name) ~ "Junior",
         grepl("Jack  Arend", Name) ~ "Junior",
         grepl("Kyle  Carter", Name) ~ "Junior",
         grepl("Justin  White", Name) ~ "Senior",
         grepl("Bryan  Gotti", Name) ~ "First-Year",
         grepl("Jon  Lindgren", Name) ~ "Junior",
         grepl("Antonio  Jareno", Name) ~ "First-Year",
         grepl("Dan  Trulli", Name) ~ "Senior",
         grepl("Andrew  Chi", Name) ~ "First-Year",
         grepl("Will  Sylvia", Name) ~ "Junior",
         grepl("Pat  Beaton", Name) ~ "Junior",
         grepl("Noah  Loughlin", Name) ~ "Sophomore",
         grepl("Kevin  Lucey", Name) ~ "First-Year",
         grepl("Colin  Coyne", Name) ~ "Sophomore",
         grepl("Leo  Watson", Name) ~ "Junior",
         grepl("Dan  Brady", Name) ~ "Sophomore",
         grepl("Chris  Capo", Name) ~ "First-Year",
         grepl("Rob  Matson", Name) ~ "Sophomore",
         TRUE ~ as.character(Year)))

#Add in Position to be used in Shiny App
nescacwar2 <- nescacwar2 %>%
  mutate(Position= Pos.x) %>%
  mutate(Position= case_when(grepl("Sr", Position) ~ "Senior",
         grepl("INF", Position) ~ "Infield",
         grepl("IF", Position) ~ "Infield",
         grepl("1B", Position) ~ "Infield",
         grepl("2B", Position) ~ "Infield",
         grepl("SS", Position) ~ "Infield",
         grepl("3B", Position) ~ "Infield",
         grepl("CF", Position) ~ "Outfield",
         grepl("C", Position) ~ "Catcher",
         grepl("OF", Position) ~ "Outfield",
         grepl("UTIL", Position) ~ "Utility",
         grepl("UTL", Position) ~ "Utility",
         grepl("UT", Position) ~ "Utility",
         grepl("DH", Position) ~ "Designated Hitter",
         grepl("P", Position) ~ "Designated Hitter",
         grepl("Matt  Santos", Name) ~ "Infield",
         grepl("Ryan  Nakajima", Name) ~ "Outfield",
         grepl("Ryan  Noone", Name) ~ "Catcher",
         grepl("Nate  Bozzella", Name) ~ "Infield",
         grepl("Adam  Geibel", Name) ~ "Infield",
         grepl("Jamie  Mills", Name) ~ "Outfield",
         grepl("Kyle  Sylvester", Name) ~ "Infield",
         grepl("Spencer  Nastala", Name) ~ "Catcher",
         grepl("Ryan Young", Name) ~ "Utility",
         grepl("Matt Koperniak", Name) ~ "Outfield",
         grepl("Geoff Stillman", Name) ~ "Outfield",
         grepl("Johnny Stamatis", Name) ~ "Infield",
         grepl("Ben Reinisch", Name) ~ "Infield",
         grepl("Alex Rodriguez", Name) ~ "Catcher",
         grepl("Mike Guanci", Name) ~ "Infield",
         grepl("Robbie Cronin", Name) ~ "Infield",
         grepl("Mack Lauder", Name) ~ "Cathcer",
         grepl("Vincent Capone", Name) ~ "Infield",
         grepl("Eric Thronson", Name) ~ "Catcher",
         grepl("Tyler Buckley", Name) ~ "Utility",
         grepl("Brett Stevenson", Name) ~ "Outfield",
         grepl("Ian Steckel", Name) ~ "Outfield",
         grepl("Gianni Valentini", Name) ~ "Infield",
         grepl("Jimmy Edwards", Name) ~ "Infield",
         grepl("Joseph Celio", Name) ~ "Outfield",
         grepl("Kevin Bomba", Name) ~ "Outfield",
         grepl("Ian Maldonado", Name) ~ "Infield",
         grepl("Alex Steele", Name) ~ "Infield",
         grepl("Cameron Crowley", Name) ~ "Infield",
         TRUE ~ as.character(Position)))


  
  
write.csv(nescacwar2, file= "nescacwar2.csv")

  
save(nescacwar2, file="nescacwar2.Rda")
```


## References

The following sources were used to find the necessary formuals to calculate fOWar, wOBA, wRAA, wSB, and wRC.

fOWar:https://library.fangraphs.com/war/war-position-players/D

wOBA: https://library.fangraphs.com/offense/woba/

Deriving wOBA: https://library.fangraphs.com/the-beginners-guide-to-deriving-woba/

wSB: https://library.fangraphs.com/offense/wsb/

wRAA: https://library.fangraphs.com/offense/wraa/

wRC: https://library.fangraphs.com/offense/wrc/

Runs Per Win: https://library.fangraphs.com/misc/war/converting-runs-to-wins/

MLB wOBA and other constants: https://www.fangraphs.com/guts.aspx?type=cn